---
title: "Interval-censored PAMMs: Monte Carlo benchmark"
date: today
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 10
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(tibble)
})
bench_dir <- if (file.exists("config.R")) "." else "attic/ic-benchmark"
res <- readRDS(file.path(bench_dir, "results", "summary.rds"))
tl <- res$term_level
pw <- res$pointwise |>
  left_join(
    distinct(tl, cell_id, arm, baseline, effect, rate, n, mean_int),
    by = "cell_id"
  )
paired <- res$paired
theme_set(theme_bw(base_size = 12))
cols <- c(
  oracle = "#1b9e77", mi = "#1f78b4", midpoint = "#e31a1c",
  ic_par = "#ff7f00", ic_sp = "#6a3d9a", turnbull = "#b15928"
)
meth_lvls <- names(cols)
tl <- tl |> mutate(method = factor(method, levels = meth_lvls))
pw <- pw |> mutate(method = factor(method, levels = meth_lvls))
n_rep_prod <- max(tl$n_rep)
```

All prose numbers in this report are computed from `results/summary.rds`
(`r res$n_raw_rows` raw result rows, up to `r n_rep_prod` replications per
cell). Monte Carlo SEs for grid-averaged coverage are clustered at the
replication level. Cells where a method is misspecified by design are flagged
with open symbols.

## Failure rates

```{r failures, results = "asis"}
fl <- res$failures |> filter(n_fail > 0)
if (nrow(fl) == 0) {
  cat("No method failures occurred in any cell.\n")
} else {
  cat(sprintf(
    "%d (cell, method) combinations had failures; max failure rate %.1f%% (%s, %s).\n\n",
    nrow(fl), 100 * max(fl$fail_rate),
    fl$cell_id[which.max(fl$fail_rate)], fl$method[which.max(fl$fail_rate)]
  ))
  knitr::kable(fl, digits = 3)
}
if (nrow(res$missing) > 0) {
  cat(sprintf("\n**WARNING: %d missing (cell, rep) combinations.**\n",
              nrow(res$missing)))
}
```

## Fig 1 (headline): coverage of S(t|x) and interval-average hazard

```{r fig1, fig-height: 7}
d1 <- tl |>
  filter(arm == "core", estimand %in% c("surv", "hazard"), !is.na(coverage))
ggplot(d1, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  geom_line(alpha = .5) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage (grid-avg)",
    colour = NULL, shape = "misspecified",
    title = "Coverage of S(t|x) and interval-average hazard, core factorial",
    subtitle = "pointranges: +-1 MC SE (rep-clustered)"
  )
```

```{r fig1-prose, results = "asis"}
mi_s <- d1 |> filter(method == "mi", estimand == "surv")
md_s <- d1 |> filter(method == "midpoint", estimand == "surv")
cat(sprintf(
  "MI survival coverage across core cells: %.3f-%.3f (median %.3f). Midpoint: %.3f-%.3f (median %.3f).\n",
  min(mi_s$coverage), max(mi_s$coverage), median(mi_s$coverage),
  min(md_s$coverage), max(md_s$coverage), median(md_s$coverage)
))
```

## Fig 2: beta / logHR(t) coverage + paired MI−midpoint differences

```{r fig2}
d2 <- tl |>
  filter(arm == "core", estimand %in% c("beta", "logHR"), !is.na(coverage))
ggplot(d2, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Covariate-effect coverage (beta in PH cells, logHR(t) grid-avg)",
    subtitle = "open symbols: misspecified by design (constant-beta methods vs beta(t))"
  )
```

```{r fig2b, fig-height: 4}
ggplot(
  paired |> filter(arm == "core"),
  aes(factor(rate, levels = c(1.5, 0.6, 0.3)), diff, colour = estimand)
) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(ymin = diff - diff_mcse, ymax = diff + diff_mcse),
    position = position_dodge(width = 0.4)
  ) +
  facet_grid(n ~ baseline + effect) +
  labs(
    x = "inspection rate (dense -> sparse)", y = "coverage(MI) - coverage(midpoint)",
    title = "Paired MI - midpoint coverage differences (same data per rep)",
    subtitle = "pointranges: +-1 paired MC SE"
  )
```

## Fig 3: point accuracy (RMSE, incl. Turnbull)

```{r fig3}
d3 <- tl |> filter(arm == "core", estimand == "surv")
base_rmse <- d3 |>
  filter(method == "oracle") |>
  select(cell_id, rmse_oracle = rmse)
d3 <- left_join(d3, base_rmse, by = "cell_id") |>
  mutate(rrmse = rmse / rmse_oracle)
ggplot(d3, aes(mean_int, rrmse, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  scale_y_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "mean inspection-interval length", y = "RMSE relative to oracle (log)",
    title = "S(t|x) point-estimate accuracy relative to the oracle"
  )
```

```{r fig3-prose, results = "asis"}
tb <- tl |> filter(method == "turnbull", estimand == "surv", arm == "core")
if (nrow(tb)) {
  rr <- d3 |>
    filter(method %in% c("mi", "midpoint", "turnbull")) |>
    group_by(method) |>
    summarise(med_rrmse = median(rrmse, na.rm = TRUE), .groups = "drop")
  n_tb_beats_mi <- d3 |>
    select(cell_id, method, rrmse) |>
    pivot_wider(names_from = method, values_from = rrmse) |>
    summarise(v = sum(turnbull < mi, na.rm = TRUE), n = sum(!is.na(turnbull)))
  cat(sprintf(
    "Median RRMSE vs oracle: MI %.2f, midpoint %.2f, Turnbull %.2f; Turnbull beats MI in %d/%d core cells (it is a step-function NPMLE without smoothing). Turnbull mean equivalence-class half-gap (scoring ambiguity) ranges %.3f-%.3f across core cells.\n",
    rr$med_rrmse[rr$method == "mi"], rr$med_rrmse[rr$method == "midpoint"],
    rr$med_rrmse[rr$method == "turnbull"],
    n_tb_beats_mi$v, n_tb_beats_mi$n,
    min(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE),
    max(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE)
  ))
}
```

## Fig 4: CI width and avSE/empSE (the mechanism)

```{r fig4}
d4 <- tl |>
  filter(arm == "core", estimand == "surv",
         method %in% c("mi", "midpoint", "oracle", "ic_par")) |>
  pivot_longer(c(median_width, se_ratio), names_to = "what")
ggplot(d4, aes(mean_int, value, colour = method)) +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(what ~ baseline + effect + n, scales = "free_y") +
  scale_colour_manual(values = cols) +
  labs(
    x = "mean inspection-interval length", y = NULL,
    title = "S(t|x): median CI width and avSE/empSE ratio (median over grid)",
    subtitle = "se_ratio < 1: model-based SE understates sampling variability"
  )
```

## Fig 5: time-varying effect recovery

```{r fig5}
lf_file <- file.path(bench_dir, "results", "least-false.rds")
d5 <- pw |>
  filter(effect == "tv", estimand == "logHR", n == 300,
         method %in% c("mi", "midpoint", "oracle", "ic_par"))
if (nrow(d5) > 0) {
  lf_lines <- if (file.exists(lf_file)) {
    readRDS(lf_file) |>
      distinct(baseline, lf_cox) |>
      semi_join(d5, by = "baseline")
  } else {
    NULL
  }
  p5 <- ggplot(d5, aes(t, mean_est, colour = method)) +
    geom_line(aes(y = truth), colour = "black", linewidth = 0.9) +
    geom_pointrange(
      aes(ymin = mean_est - bias_mcse, ymax = mean_est + bias_mcse),
      position = position_dodge(width = 0.3)
    ) +
    geom_line(alpha = 0.5, position = position_dodge(width = 0.3)) +
    facet_grid(baseline ~ rate) +
    scale_colour_manual(values = cols) +
    labs(
      x = "t", y = "logHR(t)",
      title = "TV cells: mean estimated logHR(t) vs truth (black), n = 300",
      subtitle = "dashed grey: least-false constant (Cox projection) -- ic_par's actual target"
    )
  if (!is.null(lf_lines)) {
    p5 <- p5 +
      geom_hline(
        data = lf_lines, aes(yintercept = lf_cox),
        linetype = 2, colour = "grey50"
      )
  }
  print(p5)
} else {
  cat("(no TV-cell data in this run)")
}
```

```{r fig5-prose, results = "asis"}
if (file.exists(lf_file)) {
  lf <- readRDS(lf_file) |> distinct(baseline, lf_cox, lf_weibull)
  for (i in seq_len(nrow(lf))) {
    cat(sprintf(
      "- %s baseline: least-false constant logHR (exact-time convention): Cox %.3f, Weibull %.3f\n",
      lf$baseline[i], lf$lf_cox[i], lf$lf_weibull[i]
    ))
  }
} else {
  cat("Least-false projections not yet computed (results/least-false.rds missing).\n")
}
```

## Fig 6: competing risks, smooth f(x), m-sensitivity

```{r fig6a}
d6 <- tl |> filter(arm %in% c("cr", "cr_unknown"), !is.na(coverage))
ggplot(d6, aes(mean_int, coverage, colour = method, shape = arm)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.1)) +
  facet_wrap(~estimand, nrow = 1) +
  scale_colour_manual(values = cols) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Competing-risks arm (triangle: 30% unknown causes)"
  )
```

```{r fig6b, fig-height: 4}
d6b <- tl |> filter(arm == "smooth", estimand == "f_x")
if (nrow(d6b)) {
  ggplot(d6b, aes(factor(rate), coverage, colour = method)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    scale_colour_manual(values = cols) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = "inspection rate", y = "95% CI coverage",
      title = "Smooth f(x) coverage (spoke S, grid-avg over x)"
    )
}
```

```{r fig6c, fig-height: 4}
d6c <- tl |>
  filter(arm == "mimp" | cell_id == "core-peaked-ph-random-r0.6-n300-m10",
         method == "mi", estimand %in% c("surv", "hazard", "beta"))
if (nrow(d6c)) {
  ggplot(d6c, aes(factor(m), coverage, colour = estimand)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    coord_cartesian(ylim = c(0.7, 1)) +
    labs(
      x = "number of imputations m", y = "95% CI coverage",
      title = "m-sensitivity (spoke I; peaked, PH, rate 0.6, n = 300)"
    )
}
```

## Term-level summary tables

```{r tables, results = "asis"}
for (es in unique(tl$estimand)) {
  cat(sprintf("\n### %s\n\n", es))
  print(knitr::kable(
    tl |>
      filter(estimand == es) |>
      mutate(
        coverage = sprintf("%.3f (%.3f)", coverage, coverage_mcse),
        bias = sprintf("%.4f (%.4f)", bias, bias_mcse),
        rmse = sprintf("%.4f (%.4f)", rmse, rmse_mcse)
      ) |>
      select(cell_id, method, n_rep, coverage, be_coverage, bias, rmse,
             se_ratio, median_width, misspecified),
    digits = 3
  ))
}
```

## Session / run metadata

```{r meta, results = "asis"}
cat(sprintf("- raw rows: %d\n", res$n_raw_rows))
cat(sprintf("- max reps/cell: %d\n", n_rep_prod))
cat(sprintf("- median fit times by method (s): %s\n",
  tl |> group_by(method) |> summarise(t = median(mean_fit_time, na.rm = TRUE)) |>
    (\(d) paste(sprintf("%s %.1f", d$method, d$t), collapse = ", "))()
))
```
