---
title: "Interval-censored PAMMs: Monte Carlo benchmark"
date: today
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 10
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(tibble)
})
bench_dir <- if (file.exists("config.R")) "." else "attic/ic-benchmark"
res <- readRDS(file.path(bench_dir, "results", "summary.rds"))
tl <- res$term_level
pw <- res$pointwise |>
  left_join(
    distinct(tl, cell_id, arm, baseline, effect, rate, n, mean_int),
    by = "cell_id"
  )
paired <- res$paired
theme_set(theme_bw(base_size = 12))
cols <- c(
  oracle = "#1b9e77", mi = "#1f78b4", midpoint = "#e31a1c",
  ic_par = "#ff7f00", ic_sp = "#6a3d9a", turnbull = "#b15928"
)
meth_lvls <- names(cols)
tl <- tl |> mutate(method = factor(method, levels = meth_lvls))
pw <- pw |> mutate(method = factor(method, levels = meth_lvls))
n_rep_prod <- max(tl$n_rep)
```

All prose numbers in this report are computed from `results/summary.rds`
(`r res$n_raw_rows` raw result rows, up to `r n_rep_prod` replications per
cell). Monte Carlo SEs for grid-averaged coverage are clustered at the
replication level. Cells where a method is misspecified by design are flagged
with open symbols.

# Design

## Aims

This benchmark validates and compares the interval-censored (IC) survival
workflow added to `pammtools`: `pamm_ic()` / `pamm_ic_cr()` fit piecewise
exponential additive mixed models (PAMMs) to interval-censored data by
multiple imputation — exact event times are drawn from a conditional
piecewise-exponential hazard initialized from a midpoint-imputed fit
(one-step MI), a PAMM is fit per imputation, and inference is pooled via
Rubin's rules. The study follows the ADEMP framework (Morris, White &
Crowther 2019, *Stat Med*).

- **A1 (validation):**  
does `pamm_ic` MI inference attain nominal 95%
  coverage and low bias for hazard, survival, covariate effects, time-varying
  effects and competing-risks CIFs, across inspection sparsity, inspection
  mechanisms and baseline-hazard shapes?
- **A2 (comparison):**  
how does it compare against established external IC
  methods — `icenReg::ic_par` (parametric Weibull PH), `icenReg::ic_sp`
  (semiparametric PH) and the Turnbull NPMLE (`icenReg::ic_np`) — including
  cells where the parametric comparator is **misspecified by design**? The
  headline misspecification exhibit is `ic_par`'s S(t|x) behavior under the
  non-Weibull "peaked" baseline; PH-only comparators in time-varying-effect
  cells are reported as *estimand mismatch* (they target a least-false
  constant logHR), never as headline "coverage failure".

The design was frozen pre-production at Gate R2 (three external review gates:
code, pilot, results; see the decision log in `README.md`). The earlier
200-rep prototype is `attic/ic-simulation-study.R`.

## Methods compared

| Method | Fit | Confidence intervals | Role |
|---|---|---|---|
| `oracle` | PAMM on the **exact** (uncensored) event times, analysis grid CUT | simulation CIs, 500 posterior draws | upper reference bound: best achievable on this grid; isolates the cost of interval censoring |
| `mi` | `pamm_ic` / `pamm_ic_cr`: m = 10 imputations (5/20 in spoke I), PAMM per imputation, Rubin-pooled | pooled posterior draws (500 total); scalar β via Rubin's rules with Barnard–Rubin small-sample df | the method under test |
| `midpoint` | single PAMM on midpoint-imputed times ((L + min(R, horizon))/2; right-censored at L if R = ∞) | simulation CIs, 500 draws | naive single-imputation comparator: ignores imputation uncertainty |
| `ic_par` | `icenReg::ic_par`, Weibull PH | Wald z for β; S/hazard via 500 normal coefficient draws pushed through the closed-form Weibull | external parametric comparator; **misspecified by design** under the peaked baseline and in TV cells |
| `ic_sp` | `icenReg::ic_sp`, semiparametric PH (NPMLE baseline) | β: bootstrap percentile (500 resamples) in 4 designated cells, point-only elsewhere; S(t\|x) point-only (the bootstrap covers coefficients, not the NPMLE baseline) | external semiparametric comparator |
| `turnbull` | `icenReg::ic_np` stratified by x (Turnbull NPMLE) | none (point accuracy only) | nonparametric accuracy benchmark for S(t\|x) in the simple cases |

All simulation-based CIs use the same number of draws (nsim = 500, equalized
so coverage differences are not tail-order-statistic artifacts; a 20-rep LRZ
sensitivity run over nsim ∈ {200, 500, 1000} found coverage differences well
below MC SE — decision log, 2026-06-10).
Within each Turnbull equivalence class (where the NPMLE is flat or
undefined), S is scored at the midpoint of the upper/lower envelope and the
half-gap is recorded as `ambiguity`. Note one deliberate construction
difference: MI β CIs are Barnard–Rubin t, midpoint/oracle/ic_par β CIs are
Wald z. External methods apply only in the single-event, linear-effect cells;
the smooth-f(x) and competing-risks arms compare internal methods only
(mi/midpoint/oracle).

Model formulas (`mgcv` syntax): 

- PH cells `~ s(tend, k = 10) + x`
- TV cells add `+ s(tend, by = x_num, k = 10)`
- spoke S `~ s(tend, k = 10) + s(x, k = 10)`
- CR arm `~ s(tend, by = cause, k = 10) + cause + cause:x`.

## Data-generating processes and design

Data are simulated with `pammtools::sim_pexp()` on a fine grid
(FINE = `seq(0, 10, 0.05)`) and interval-censored with
`pammtools::add_inspections()`. All methods analyze the same frozen grid
(CUT = `seq(0, 10, 0.5)`); estimands are evaluated at t ∈ {1, 2, 3, 5, 7} (TEVAL).

**Hazards.** 

Log-baseline shapes: 
 
- *const* −1.5 (Weibull-compatible; `ic_par`
correctly specified)
- *peaked* −3.2 + 1.6·exp(−(t−3)²/2) (non-monotone; no
Weibull can match it)
- *rising* −3.7 + 0.12t (spoke R). 

Covariate effects:

- *ph* β·x with β = 0.6 and binary x (p = 0.5 — chosen so the stratified
Turnbull NPMLE is directly comparable)
- *tv* β(t)·x with
β(t) = 1.2·exp(−0.3t)
- *smooth* f(x) = 0.8·sin(πx) with x ~ U(−1, 1). 

The CR arm uses two independent latent cause-specific hazards (cause 1 peaked with
β₁ = 0.7, cause 2 rising with β₂ = −0.5, x ~ U(−1, 1)); the `cr_unknown` cell
additionally masks 30% of observed event causes to exercise `pamm_ic_cr`'s cause-imputation branch.

**Inspection mechanisms.** 

- *random*: per-subject inspection grids with
Exp(rate) gaps, rate ∈ {1.5, 0.6, 0.3} — mean gap 1/rate ≈ 0.67 / 1.67 / 3.33,
i.e. dense → sparse interval censoring. 
- *fixed*: one shared deterministic schedule for all subjects (dense gap 2/3, sparse gap 10/3). 
- *mixed*: the fixed schedule shifted per subject by U(0, gap). 

Every inspection grid additionally contains an **end-of-study examination at the horizon** (`terminal_exam = TRUE`
in `add_inspections`), so the event time is always bracketed by the nearest
inspections (L, R] with R ≤ 10, and only subjects event-free at the horizon are
right-censored. (Mixing conventions — exact exit times for survivors but open
intervals for undetected events — would make the right-censoring informative;
the fraction of events bracketed only by the terminal exam is recorded as `prop_terminal_only`.)  

**Caveat (fixed schedule):** with one shared grid, S(t) is
nonparametrically identified *only at the schedule points*; elsewhere all
methods score model-based interpolation, and sparse-grid hazard estimands can
be degenerate. These cells are robustness exhibits, never headlined alongside
the random-inspection cells.

**Truth computation.** `sim_pexp()` draws from a *left-endpoint
piecewise-constant* hazard on FINE, not from the smooth formula. All truths
(S, interval-average hazard, logHR, CIF) are therefore computed **exactly**
from those same FINE rate vectors via cumulative sums (`dgp.R`) — truth and
simulator derive from one definition and cannot diverge. Phase 0 verified
this against a large-sample Kaplan–Meier (max |ΔS| ≤ 0.0016) and an
independent `survival::survfit` Turnbull cross-check.

**Design.** A core factorial plus targeted spokes; the cells present in this
run:

```{r design-table}
design <- tl |>
  distinct(cell_id, arm, baseline, effect, mechanism, rate, n, m) |>
  group_by(arm) |>
  summarise(
    cells = n(),
    baseline = paste(sort(unique(baseline)), collapse = ", "),
    effect = paste(sort(unique(effect)), collapse = ", "),
    mechanism = paste(sort(unique(mechanism)), collapse = ", "),
    rate = paste(sort(unique(rate), decreasing = TRUE), collapse = ", "),
    n = paste(sort(unique(n)), collapse = ", "),
    m = paste(sort(unique(m)), collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(match(arm, c(
    "core", "mech", "rising", "smooth", "mimp", "cr", "cr_unknown"
  )))
knitr::kable(design)
```

The core arm crosses baseline × effect × inspection rate × n under random
inspections; the spokes vary one factor each off the (peaked, ph, n = 300)
reference cell: inspection *mech*anism, *rising* baseline, *smooth* f(x),
number of imputations (*mimp*), and competing risks.  

Acknowledged gaps:  
mechanism × n and mechanism × effect interactions are untested, so spoke
conclusions are conditional on the reference cell.

## Estimands

All estimands are aligned exactly to the piecewise-exponential DGP:

- **S(t|x)** at TEVAL × x ∈ {0, 1} — well-defined for every method, hence the
  headline estimand for cross-method comparison.
- **Hazard** as the *interval average* h̄ⱼ = (H(tⱼ) − H(tⱼ₋₁)) / 0.5 on the CUT
  interval ending at each TEVAL. Rationale: a PAMM estimates piecewise-constant
  hazards on CUT; scoring it against the smooth endpoint hazard h(tⱼ) would
  build up to ~25% "bias" on the peak flanks for *all* methods — a
  discretization artifact, not a method property. `ic_par` is scored the same
  way via closed-form cumulative-hazard differences.
- **logHR(t)** = log h̄ⱼ(1) − log h̄ⱼ(0); constant = β under PH. In TV cells the
  PH-only comparators are evaluated against the *least-false constant logHR*
  (Cox projection from one large-n exact-time fit; overlay in Fig 5) rather
  than as covering β(t).
- **β** (scalar, PH cells) 
- **f(x)** on an x-grid, with estimate and truth both
  centered to mean zero over the grid (spoke S)
- **CIF₁/CIF₂(t)** at x = 0 (CR arm).

## Performance measures and Monte Carlo precision

Per Morris et al. (2019): bias, empirical SE, RMSE, 95% CI coverage,
bias-eliminated coverage (CI shifted to cover the mean estimate — separates
"CIs too narrow" from "point estimate biased"), CI width, the avSE/empSE
ratio (model-based vs empirical SE; the most diagnostic quantity behind
coverage failures), failure rates and fit times — each with Monte Carlo SEs.
Two aggregation details matter:

- avSE/empSE for grid-valued estimands is computed *within* each (t, x) grid
  point and then median-summarized; pooling errors across the grid mixes
  cross-point bias heterogeneity into "empSE" and invalidates the ratio.
- CI widths are summarized by medians and q90 with a flag for extreme widths,
  since single pathological replications (vacuous CIs in weakly identified
  cells) make raw means meaningless.

n_sim = 500 reps/cell gives a coverage MC SE of ≈ 1.0% at 95% nominal. The
headline MI-vs-midpoint contrasts therefore use **paired differences** (both
methods see the same data in each rep) with paired MC SEs, which are
substantially more precise than per-method coverage; term-level
(grid-averaged) coverage MC SEs are clustered at the replication level.

# Results

## Failure rates

```{r failures, results = "asis"}
fl <- res$failures |> filter(n_fail > 0)
if (nrow(fl) == 0) {
  cat("No method failures occurred in any cell.\n")
} else {
  cat(sprintf(
    "%d (cell, method) combinations had failures; max failure rate %.1f%% (%s, %s).\n\n",
    nrow(fl), 100 * max(fl$fail_rate),
    fl$cell_id[which.max(fl$fail_rate)], fl$method[which.max(fl$fail_rate)]
  ))
  knitr::kable(fl |> select(-warn_rate, -first_warning), digits = 3)
}
if (nrow(res$missing) > 0) {
  cat(sprintf("\n**WARNING: %d missing (cell, rep) combinations.**\n",
              nrow(res$missing)))
}
wl <- res$failures |> filter(warn_rate > 0.10)
if (nrow(wl) > 0) {
  cat(sprintf(
    "\nFailures are not the whole story: %d (cell, method) combinations had fitter warnings in >10%% of reps (these reps ARE included in all summaries):\n\n",
    nrow(wl)
  ))
  print(knitr::kable(
    wl |>
      mutate(first_warning = substr(first_warning, 1, 60)) |>
      select(cell_id, method, warn_rate, first_warning),
    digits = 3
  ))
  cat(
    "\nSystematic warnings are concentrated in the fixed-schedule robustness",
    "cells, where the shared sparse inspection grid leaves hazards weakly",
    "identified between schedule points (see Design: fixed-schedule caveat).\n"
  )
}
```

## Fig 1 (headline): coverage of S(t|x) and interval-average hazard

```{r fig1, fig-height: 7}
d1 <- tl |>
  filter(arm == "core", estimand %in% c("surv", "hazard"), !is.na(coverage))
ggplot(d1, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  geom_line(alpha = .5) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage (grid-avg)",
    colour = NULL, shape = "misspecified",
    title = "Coverage of S(t|x) and interval-average hazard, core factorial",
    subtitle = "pointranges: +-1 MC SE (rep-clustered)"
  )
```

```{r fig1-prose, results = "asis"}
mi_s <- d1 |> filter(method == "mi", estimand == "surv")
md_s <- d1 |> filter(method == "midpoint", estimand == "surv")
cat(sprintf(
  "MI survival coverage across core cells: %.3f-%.3f (median %.3f). Midpoint: %.3f-%.3f (median %.3f).\n",
  min(mi_s$coverage), max(mi_s$coverage), median(mi_s$coverage),
  min(md_s$coverage), max(md_s$coverage), median(md_s$coverage)
))
mi_d <- d1 |> filter(method == "mi", rate == 1.5)
for (es in c("surv", "hazard")) {
  dd <- mi_d |> filter(estimand == es)
  worst <- dd |> slice_min(coverage, n = 1)
  cat(sprintf(
    "Under dense inspection (rate 1.5), MI %s coverage is %.3f-%.3f; the worst cell (%s: %.3f, MC SE %.3f) has bias-eliminated coverage %.3f, i.e. any shortfall is point-estimate bias, not SE failure.\n",
    es, min(dd$coverage), max(dd$coverage),
    worst$cell_id, worst$coverage, worst$coverage_mcse, worst$be_coverage
  ))
}
```

```{r fig1-sparse-prose, results = "asis"}
# headline sparse-cell story (computed; required at Gate R3-final)
pto <- mean(tl$prop_terminal_only[tl$arm == "core" & tl$rate == 0.3],
            na.rm = TRUE)
grad <- tl |>
  filter(arm == "core", rate == 0.3, effect == "ph", estimand == "surv",
         method == "mi") |>
  arrange(baseline, n)
gtxt <- paste(
  sprintf("%s n=%d: %.3f (BE %.3f)",
          grad$baseline, grad$n, grad$coverage, grad$be_coverage),
  collapse = "; "
)
p1 <- pw |>
  filter(cell_id == "core-const-ph-random-r0.3-n1000-m10",
         estimand == "surv", t == 1, x == 0)
mi1 <- p1 |> filter(method == "mi")
tb1 <- p1 |> filter(method == "turnbull")
md_sp <- tl |>
  filter(arm == "core", rate == 0.3, estimand == "surv", method == "midpoint")
cat(sprintf(
  "**Sparse inspection (rate 0.3; %.1f%% of events bracketed only by the end-of-study exam).** MI under-covers S(t|x), and the shortfall grows with n at fixed bias (grid-averaged coverage, BE = bias-eliminated: %s) -- the signature of point-estimate bias with honest SEs. The bias is genuine one-step-MI bias concentrated at early-to-mid t: in const-ph/n1000, S(1|0) bias is %+.3f (coverage %.3f), while the Turnbull NPMLE on the *same data* recovers S(1|0) with bias %+.3f. The information is in the data; the single midpoint-initialized imputation model leaves it on the table. Midpoint imputation is far worse still (coverage %.3f-%.3f): events bracketed only by the terminal exam are imputed at (L+10)/2, beyond their conditional expectation under the true hazard, a structural bias that one-step MI attenuates but does not remove.\n",
  100 * pto, gtxt, mi1$bias, mi1$coverage, tb1$bias,
  min(md_sp$coverage), max(md_sp$coverage)
))

# estimand caveat: interval-average hazard at peaked baselines (oracle ceiling)
hz <- tl |>
  filter(arm == "core", baseline == "peaked", effect == "ph", n == 1000,
         estimand == "hazard", method == "mi") |>
  arrange(desc(rate))
orc <- tl |>
  filter(arm == "core", baseline == "peaked", estimand == "hazard",
         method == "oracle")
cat(sprintf(
  "\n**Estimand caveat: interval-average hazard at peaked baselines.** MI hazard coverage in peaked-ph/n1000 is %s across rates 1.5/0.6/0.3 (BE %s) -- peak-attenuation point bias, not SE failure. The oracle itself attains only %.3f-%.3f on this estimand across peaked cells: a penalized-spline smoothing ceiling, not an interval-censoring phenomenon.\n",
  paste(sprintf("%.3f", hz$coverage), collapse = "/"),
  paste(sprintf("%.3f", hz$be_coverage), collapse = "/"),
  min(orc$coverage), max(orc$coverage)
))
```

## Fig 2: beta / logHR(t) coverage + paired MI−midpoint differences

```{r fig2}
d2 <- tl |>
  filter(arm == "core", estimand %in% c("beta", "logHR"), !is.na(coverage))
ggplot(d2, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Covariate-effect coverage (beta in PH cells, logHR(t) grid-avg)",
    subtitle = "open symbols: misspecified by design (constant-beta methods vs beta(t))"
  )
```

```{r fig2b, fig-height: 4}
ggplot(
  paired |> filter(arm == "core"),
  aes(factor(rate, levels = c(1.5, 0.6, 0.3)), diff, colour = estimand)
) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(ymin = diff - diff_mcse, ymax = diff + diff_mcse),
    position = position_dodge(width = 0.4)
  ) +
  facet_grid(n ~ baseline + effect) +
  labs(
    x = "inspection rate (dense -> sparse)", y = "coverage(MI) - coverage(midpoint)",
    title = "Paired MI - midpoint coverage differences (same data per rep)",
    subtitle = "pointranges: +-1 paired MC SE"
  )
```

```{r fig2b-prose, results = "asis"}
pd <- paired |>
  filter(arm == "core") |>
  group_by(grp = ifelse(estimand %in% c("surv", "hazard"), "S/hazard", "beta/logHR")) |>
  summarise(
    lo = min(diff), hi = max(diff),
    n_pos = sum(diff > 2 * diff_mcse),
    n_neg = sum(diff < -2 * diff_mcse),
    n = n(), .groups = "drop"
  )
for (i in seq_len(nrow(pd))) {
  cat(sprintf(
    "- Paired MI - midpoint coverage differences, %s: range %+.3f to %+.3f; MI significantly better (>2 paired MC SE) in %d/%d contrasts, significantly worse in %d.\n",
    pd$grp[i], pd$lo[i], pd$hi[i], pd$n_pos[i], pd$n[i], pd$n_neg[i]
  ))
}
sc <- pd[pd$grp == "beta/logHR", ]
cat(sprintf(
  "\nMI's coverage advantage over midpoint imputation is strongest for S(t|x)/hazard; for the scalar effect estimands (beta, logHR), MI is significantly better in %d and significantly worse in %d of %d contrasts.\n",
  sc$n_pos, sc$n_neg, sc$n
))
bb <- tl |>
  filter(method == "mi", estimand == "beta", arm == "core") |>
  mutate(z = bias / bias_mcse) |>
  filter(abs(z) > 2)
if (nrow(bb)) {
  cat(sprintf(
    "\nMI beta bias is negligible except in %s.\n",
    paste(sprintf("%s (%+.4f, z = %.1f; coverage %.3f)",
                  bb$cell_id, bb$bias, bb$z, bb$coverage), collapse = "; ")
  ))
}
```

## Fig 3: point accuracy (RMSE, incl. Turnbull)

```{r fig3}
d3 <- tl |> filter(arm == "core", estimand == "surv")
base_rmse <- d3 |>
  filter(method == "oracle") |>
  select(cell_id, rmse_oracle = rmse)
d3 <- left_join(d3, base_rmse, by = "cell_id") |>
  mutate(rrmse = rmse / rmse_oracle)
ggplot(d3, aes(mean_int, rrmse, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  scale_y_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "mean inspection-interval length", y = "RMSE relative to oracle (log)",
    title = "S(t|x) point-estimate accuracy relative to the oracle"
  )
```

```{r fig3-prose, results = "asis"}
tb <- tl |> filter(method == "turnbull", estimand == "surv", arm == "core")
if (nrow(tb)) {
  rr <- d3 |>
    filter(method %in% c("mi", "midpoint", "turnbull")) |>
    group_by(method) |>
    summarise(med_rrmse = median(rrmse, na.rm = TRUE), .groups = "drop")
  n_tb_beats_mi <- d3 |>
    select(cell_id, method, rrmse) |>
    pivot_wider(names_from = method, values_from = rrmse) |>
    summarise(v = sum(turnbull < mi, na.rm = TRUE), n = sum(!is.na(turnbull)))
  cat(sprintf(
    "Median RRMSE vs oracle: MI %.2f, midpoint %.2f, Turnbull %.2f; Turnbull beats MI in %d/%d core cells (it is a step-function NPMLE without smoothing). Turnbull mean equivalence-class half-gap (scoring ambiguity) ranges %.3f-%.3f across core cells.\n",
    rr$med_rrmse[rr$method == "mi"], rr$med_rrmse[rr$method == "midpoint"],
    rr$med_rrmse[rr$method == "turnbull"],
    n_tb_beats_mi$v, n_tb_beats_mi$n,
    min(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE),
    max(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE)
  ))
}
lr <- tl |>
  filter(arm == "core", effect == "tv", n == 300, estimand == "logHR",
         method %in% c("mi", "midpoint")) |>
  select(cell_id, method, rmse) |>
  pivot_wider(names_from = method, values_from = rmse) |>
  mutate(rel = mi / midpoint)
cat(sprintf(
  "\nMI's coverage dominance does not extend to RMSE everywhere: for logHR(t) in the TV n=300 cells, MI RMSE is %.0f%%-%.0f%% of midpoint's (values >100%% = MI less accurate).\n",
  100 * min(lr$rel), 100 * max(lr$rel)
))
```

## Fig 4: CI width and avSE/empSE (the mechanism)

```{r fig4}
d4 <- tl |>
  filter(arm == "core", estimand == "surv",
         method %in% c("mi", "midpoint", "oracle", "ic_par")) |>
  pivot_longer(c(median_width, se_ratio), names_to = "what")
ggplot(d4, aes(mean_int, value, colour = method)) +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(what ~ baseline + effect + n, scales = "free_y") +
  scale_colour_manual(values = cols) +
  labs(
    x = "mean inspection-interval length", y = NULL,
    title = "S(t|x): median CI width and avSE/empSE ratio (median over grid)",
    subtitle = "se_ratio < 1: model-based SE understates sampling variability"
  )
```

```{r fig4-prose, results = "asis"}
f4 <- tl |>
  filter(arm == "core", estimand == "surv",
         method %in% c("oracle", "mi", "midpoint", "ic_par"))
sr <- range(f4$se_ratio)
wr <- f4 |>
  filter(method %in% c("mi", "midpoint")) |>
  select(cell_id, rate, method, median_width) |>
  pivot_wider(names_from = method, values_from = median_width) |>
  mutate(r = mi / midpoint)
cat(sprintf(
  "**Interpretation.** Model-based SEs for S(t|x) are essentially honest for every method: se_ratio spans %.2f-%.2f across all core cells and methods, so no method fails by understating its own sampling variability here -- the S(t|x) coverage failures of Fig 1 are point-estimate bias, not SE failure. (This panel analyzes S(t|x) only; the other estimands' SE calibration is wider but their failures are likewise bias-dominated -- see the ANOVA section and appendix tables.) What separates MI from midpoint mechanically is the width column: MI intervals are wider by a median factor of %.2f (median %.2f in the sparsest cells), which is exactly the propagated imputation uncertainty that midpoint discards. Midpoint's narrower intervals around biased point estimates produce its coverage collapse; MI's wider intervals buy back most -- not all -- of it.\n",
  sr[1], sr[2], median(wr$r), median(wr$r[wr$rate == 0.3])
))
```

## Fig 5: time-varying effect recovery

```{r fig5}
lf_file <- file.path(bench_dir, "results", "least-false.rds")
d5 <- pw |>
  filter(effect == "tv", estimand == "logHR", n == 300,
         method %in% c("mi", "midpoint", "oracle", "ic_par"))
if (nrow(d5) > 0) {
  lf_lines <- if (file.exists(lf_file)) {
    readRDS(lf_file) |>
      distinct(baseline, lf_cox) |>
      semi_join(d5, by = "baseline")
  } else {
    NULL
  }
  p5 <- ggplot(d5, aes(t, mean_est, colour = method)) +
    geom_line(aes(y = truth), colour = "black", linewidth = 0.9) +
    geom_pointrange(
      aes(ymin = mean_est - bias_mcse, ymax = mean_est + bias_mcse),
      position = position_dodge(width = 0.3)
    ) +
    geom_line(alpha = 0.5, position = position_dodge(width = 0.3)) +
    facet_grid(baseline ~ rate) +
    scale_colour_manual(values = cols) +
    labs(
      x = "t", y = "logHR(t)",
      title = "TV cells: mean estimated logHR(t) vs truth (black), n = 300",
      subtitle = "dashed grey: least-false constant (Cox projection) -- ic_par's actual target"
    )
  if (!is.null(lf_lines)) {
    p5 <- p5 +
      geom_hline(
        data = lf_lines, aes(yintercept = lf_cox),
        linetype = 2, colour = "grey50"
      )
  }
  print(p5)
} else {
  cat("(no TV-cell data in this run)")
}
```

```{r fig5-prose, results = "asis"}
if (file.exists(lf_file)) {
  lf <- readRDS(lf_file) |> distinct(baseline, lf_cox, lf_weibull)
  for (i in seq_len(nrow(lf))) {
    cat(sprintf(
      "- %s baseline: least-false constant logHR (exact-time convention): Cox %.3f, Weibull %.3f\n",
      lf$baseline[i], lf$lf_cox[i], lf$lf_weibull[i]
    ))
  }
  d5s <- d5 |>
    group_by(method) |>
    summarise(mad = mean(abs(mean_est - truth)), .groups = "drop")
  g5 <- function(m) d5s$mad[d5s$method == m]
  icp <- d5 |>
    filter(method == "ic_par") |>
    left_join(lf |> select(baseline, lf_cox), by = "baseline") |>
    summarise(mad_lf = mean(abs(mean_est - lf_cox)))
  cat(sprintf(
    "\n**Interpretation.** The internal methods recover the time-varying effect: mean |E(estimate) - beta(t)| over the (t, rate, baseline) grid is %.3f for the oracle, %.3f for MI and %.3f for midpoint -- interval censoring costs surprisingly little here because beta(t) is smooth and the by-spline pools over time. ic_par cannot bend: its mean distance from beta(t) is %.3f, but only %.3f from its least-false constant (dashed line) -- it estimates the Cox projection it targets, an estimand mismatch as designed, not an estimation failure.\n",
    g5("oracle"), g5("mi"), g5("midpoint"),
    g5("ic_par"), icp$mad_lf
  ))
} else {
  cat("Least-false projections not yet computed (results/least-false.rds missing).\n")
}
```

## Fig 6: competing risks, smooth f(x), m-sensitivity

```{r fig6a}
d6 <- tl |> filter(arm %in% c("cr", "cr_unknown"), !is.na(coverage))
ggplot(d6, aes(mean_int, coverage, colour = method, shape = arm)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.1)) +
  facet_wrap(~estimand, nrow = 1) +
  scale_colour_manual(values = cols) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Competing-risks arm (triangle: 30% unknown causes)"
  )
```

```{r fig6b, fig-height: 4}
d6b <- tl |> filter(arm == "smooth", estimand == "f_x")
if (nrow(d6b)) {
  ggplot(d6b, aes(factor(rate), coverage, colour = method)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    scale_colour_manual(values = cols) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = "inspection rate", y = "95% CI coverage",
      title = "Smooth f(x) coverage (spoke S, grid-avg over x)"
    )
}
```

```{r fig6c, fig-height: 4}
d6c <- tl |>
  filter(arm == "mimp" | cell_id == "core-peaked-ph-random-r0.6-n300-m10",
         method == "mi", estimand %in% c("surv", "hazard", "beta"))
if (nrow(d6c)) {
  ggplot(d6c, aes(factor(m), coverage, colour = estimand)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    coord_cartesian(ylim = c(0.7, 1)) +
    labs(
      x = "number of imputations m", y = "95% CI coverage",
      title = "m-sensitivity (spoke I; peaked, PH, rate 0.6, n = 300)"
    )
}
```

```{r fig6-prose, results = "asis"}
# CR arm
cr_mi <- tl |> filter(arm %in% c("cr", "cr_unknown"), method == "mi")
cr_cmp <- tl |>
  filter(rate == 0.6, arm %in% c("cr", "cr_unknown"), method == "mi") |>
  select(arm, estimand, coverage, coverage_mcse) |>
  pivot_wider(names_from = arm, values_from = c(coverage, coverage_mcse)) |>
  mutate(
    d = coverage_cr_unknown - coverage_cr,
    lim = 2 * sqrt(coverage_mcse_cr^2 + coverage_mcse_cr_unknown^2)
  )
cif_sp <- tl |>
  filter(arm == "cr", method == "mi", estimand == "cif1", rate == 0.3)
cat(sprintf(
  "**Competing risks (Fig 6a).** pamm_ic_cr's MI coverage spans %.3f-%.3f across CIF_1/CIF_2/beta_1/beta_2 and inspection rates (vs oracle %.3f-%.3f): the cause-specific MI pipeline broadly mirrors the single-event behaviour, including its sparse-cell bias signature (CIF_1 at rate 0.3: coverage %.3f, bias-eliminated %.3f). Masking 30%% of causes (cr_unknown, triangles) changes MI coverage by at most %+.3f vs the matched known-cause cell (all |differences| within 2 MC SE: %s) -- the cause-imputation branch neither breaks nor visibly degrades inference at this masking level.\n",
  min(cr_mi$coverage), max(cr_mi$coverage),
  min(tl$coverage[tl$arm == "cr" & tl$method == "oracle"]),
  max(tl$coverage[tl$arm == "cr" & tl$method == "oracle"]),
  cif_sp$coverage, cif_sp$be_coverage,
  cr_cmp$d[which.max(abs(cr_cmp$d))],
  ifelse(all(abs(cr_cmp$d) < cr_cmp$lim), "yes", "NO")
))

# smooth f(x)
d6b2 <- tl |> filter(arm == "smooth", estimand == "f_x")
if (nrow(d6b2)) {
  s6 <- d6b2 |>
    arrange(desc(rate)) |>
    group_by(method) |>
    summarise(txt = paste(sprintf("%.3f", coverage), collapse = "/"),
              .groups = "drop")
  cat(sprintf(
    "\n**Smooth f(x) (Fig 6b).** Coverage at rates 1.5/0.3: %s. The smooth covariate effect is essentially IC-insensitive, as anticipated in the design (it does not vary over time, so coarsening event times barely touches it): every method tracks the oracle, though all sit a few MC SE below nominal, oracle included -- a spline/smoothing property, not an interval-censoring one.\n",
    paste(sprintf("%s %s", s6$method, s6$txt), collapse = "; ")
  ))
}

# m-sensitivity
d6c2 <- tl |>
  filter(arm == "mimp" | cell_id == "core-peaked-ph-random-r0.6-n300-m10",
         method == "mi", estimand %in% c("surv", "hazard", "beta"))
if (nrow(d6c2)) {
  spread_m <- d6c2 |>
    group_by(estimand) |>
    summarise(spread = max(coverage) - min(coverage),
              # null limit for a pairwise difference of two coverages
              lim = 2 * sqrt(2) * max(coverage_mcse), .groups = "drop")
  cat(sprintf(
    "\n**Imputation count (Fig 6c).** Across m = 5/10/20 the coverage spread per estimand is %s (each below its two-MC-SE pairwise-difference limit of %s): no detectable m-sensitivity at this precision, so the package default m = 10 is comfortable and even m = 5 costs nothing measurable here.\n",
    paste(sprintf("%s %.3f", spread_m$estimand, spread_m$spread), collapse = ", "),
    paste(sprintf("%.3f", spread_m$lim), collapse = "/")
  ))
}
```

## Which design factors drive performance? (ANOVA meta-models)

To quantify *which* DGP characteristics and method choices drive performance,
we treat the core factorial as a designed experiment on the cell-level
summaries and fit ANOVA meta-models with all two-way interactions,

`y ~ (method + baseline + effect + rate + n)^2`,

for each suitable performance measure. The factorial is fully crossed and
balanced within each method subset, so the sums-of-squares decomposition is
order-invariant; we report each term's share of total SS (η², in %). One
observation = one (cell, method) summary, so "Residuals" collects all
three-way-and-higher interactions plus the Monte Carlo error of the
cell-level summaries. This is a *descriptive* variance
decomposition (no inferential F-tests are warranted at n = 1 summary per
cell × method); ic_par's designed misspecification is part of the design and
shows up in the method interactions.

```{r anova, results = "asis"}
core_a <- tl |>
  filter(arm == "core") |>
  mutate(
    method = droplevels(method),
    baseline = factor(baseline), effect = factor(effect),
    rate = factor(rate), n = factor(n)
  )
or_rmse <- core_a |>
  filter(method == "oracle", estimand == "surv") |>
  select(cell_id, rmse_o = rmse)
ci_methods <- c("oracle", "mi", "midpoint", "ic_par")

dat_list <- list(
  "coverage S(t|x)" = core_a |>
    filter(estimand == "surv", method %in% ci_methods) |>
    mutate(y = coverage),
  "coverage hazard" = core_a |>
    filter(estimand == "hazard", method %in% ci_methods) |>
    mutate(y = coverage),
  "log RRMSE S(t|x)" = core_a |>
    filter(estimand == "surv", method != "oracle") |>
    left_join(or_rmse, by = "cell_id") |>
    mutate(y = log(rmse / rmse_o)),
  "log avSE/empSE S(t|x)" = core_a |>
    filter(estimand == "surv", method %in% ci_methods) |>
    mutate(y = log(se_ratio))
)
eta2_of <- function(d) {
  d$method <- droplevels(d$method)
  ss <- summary(
    stats::aov(y ~ (method + baseline + effect + rate + n)^2, data = d)
  )[[1]]
  tibble(term = trimws(rownames(ss)), eta2 = ss[["Sum Sq"]] / sum(ss[["Sum Sq"]]))
}
etas <- map(dat_list, eta2_of)
wide <- reduce(
  imap(etas, \(e, nm) e |> mutate("{nm}" := round(100 * eta2, 1)) |>
         select(-eta2)),
  full_join, by = "term"
)
mx <- apply(wide[, -1], 1, max, na.rm = TRUE)
wide <- bind_rows(
  wide[wide$term != "Residuals", ][order(-mx[wide$term != "Residuals"]), ],
  wide[wide$term == "Residuals", ]
)
knitr::kable(
  wide,
  caption = "Share of total SS (eta-squared, %) per model term, core factorial. Coverage measures: methods with CIs (oracle/mi/midpoint/ic_par); log RRMSE: all non-oracle methods, RMSE relative to the oracle in the same cell."
)
```

```{r anova-prose, results = "asis"}
top_terms <- function(nm, k = 3) {
  etas[[nm]] |>
    filter(term != "Residuals") |>
    slice_max(eta2, n = k) |>
    (\(d) paste(sprintf("%s (%.0f%%)", d$term, 100 * d$eta2), collapse = ", "))()
}
g2 <- function(nm, term) {
  100 * etas[[nm]]$eta2[etas[[nm]]$term == term]
}
sds <- purrr::map_dbl(dat_list, \(d) stats::sd(d$y))
cat(sprintf(
  "Total response SDs being decomposed: %s. (Shares of a small total are still small effects: the avSE/empSE column partitions practically negligible variation.)\n\n",
  paste(sprintf("%s %.3f", names(sds), sds), collapse = "; ")
))
for (nm in names(etas)) {
  cat(sprintf("- **%s**: dominated by %s.\n", nm, top_terms(nm)))
}
mi_md_sr <- tl |>
  filter(arm == "core", estimand == "surv",
         method %in% c("oracle", "mi", "midpoint", "ic_par")) |>
  group_by(grp = method %in% c("mi", "ic_par")) |>
  summarise(med = median(se_ratio), .groups = "drop")
cat(sprintf(
  "\n**Interpretation.** *Coverage* differences are overwhelmingly about *which method meets which inspection regime and baseline shape*: for S(t|x), the method main effect plus its rate and baseline interactions carry %.0f%% of the variation (method %.0f%%, method:baseline %.0f%%, method:rate %.0f%%), while all purely-DGP terms together explain only %.0f%% -- a good method is simply robust where a bad one is not. method:baseline is the parametric-misspecification axis (ic_par's Weibull vs the peaked baseline); method:rate is the MI-vs-midpoint sparsity-robustness story of Fig 1. Hazard coverage adds a genuine baseline main effect (%.0f%%): the peaked shape is harder for *everyone*, including the oracle (the smoothing ceiling). *Point accuracy* (log RRMSE vs oracle) is structured differently: inspection rate itself is the largest single term (%.0f%%) -- coarser information costs every method accuracy relative to the oracle -- followed by the misspecification axis method:baseline (%.0f%%), while the method main effect is modest (%.0f%%): apart from parametric misspecification, the IC approaches are closer in accuracy than their coverage gaps suggest. The *avSE/empSE* column has a method term (%.0f%%) but decomposes an SD of just %.3f: a tiny systematic level shift (MI/ic_par median %.2f, midpoint/oracle %.2f) of no practical consequence. S(t|x) SE calibration is thus sound for every method, so its coverage failures are point-estimate-bias failures; bias-eliminated coverage >= %.2f across every core CI-method summary says the same of the dominant failures for the other estimands. (The RRMSE column's method set additionally contains the point-only ic_sp and Turnbull.) Finally, n terms stay <= %.0f%% in every column: with bias-driven failures, more data does not help; it sharpens the interval around the wrong value.\n",
  g2("coverage S(t|x)", "method") + g2("coverage S(t|x)", "method:rate") +
    g2("coverage S(t|x)", "method:baseline"),
  g2("coverage S(t|x)", "method"),
  g2("coverage S(t|x)", "method:baseline"),
  g2("coverage S(t|x)", "method:rate"),
  sum(100 * etas[["coverage S(t|x)"]]$eta2[
    !grepl("method|Residuals", etas[["coverage S(t|x)"]]$term)
  ]),
  g2("coverage hazard", "baseline"),
  g2("log RRMSE S(t|x)", "rate"),
  g2("log RRMSE S(t|x)", "method:baseline"),
  g2("log RRMSE S(t|x)", "method"),
  g2("log avSE/empSE S(t|x)", "method"),
  sds[["log avSE/empSE S(t|x)"]],
  mi_md_sr$med[mi_md_sr$grp], mi_md_sr$med[!mi_md_sr$grp],
  min(tl$be_coverage[tl$arm == "core" & tl$method %in% ci_methods],
      na.rm = TRUE),
  max(purrr::map_dbl(names(etas), \(nm) max(g2(nm, "n"), g2(nm, "method:n"))))
))
```

# Conclusions

```{r conclusions, results = "asis"}
# all numbers recomputed here so this section is self-contained
mi_dense <- tl |>
  filter(arm == "core", rate == 1.5, method == "mi",
         estimand %in% c("surv", "beta", "logHR"))
ph_mid <- tl |>
  filter(arm == "core", rate == 0.6, method == "mi", estimand == "surv")
pd_sp <- paired |>
  filter(arm == "core", rate == 0.3, estimand %in% c("surv", "hazard"))
mi_sp_n <- tl |>
  filter(arm == "core", rate == 0.3, effect == "ph", estimand == "surv",
         method == "mi")
icp_const <- tl |>
  filter(arm == "core", baseline == "const", effect == "ph",
         method == "ic_par", estimand == "surv")
icp_peak <- tl |>
  filter(arm == "core", baseline == "peaked", method == "ic_par",
         estimand == "surv")
cr_rng <- range(
  tl$coverage[tl$arm %in% c("cr", "cr_unknown") & tl$method == "mi"],
  na.rm = TRUE
)
rr_c <- tl |>
  filter(arm == "core", estimand == "surv") |>
  left_join(
    tl |>
      filter(arm == "core", estimand == "surv", method == "oracle") |>
      select(cell_id, rmse_o = rmse),
    by = "cell_id"
  ) |>
  mutate(rrmse = rmse / rmse_o)
rr_med <- rr_c |>
  filter(method %in% c("mi", "midpoint", "turnbull")) |>
  group_by(method) |>
  summarise(m = median(rrmse), .groups = "drop")
rrm <- function(x) rr_med$m[rr_med$method == x]
tb_v_mi <- rr_c |>
  select(cell_id, method, rrmse) |>
  pivot_wider(names_from = method, values_from = rrmse) |>
  summarise(v = sum(turnbull < mi, na.rm = TRUE), n = sum(!is.na(turnbull)))
cat(sprintf(
"1. **The MI workflow is valid where its imputation model has enough to work with.** Under dense inspection (mean gap 0.67) MI attains %.3f-%.3f coverage across survival, beta and logHR(t) in every core cell; at moderate sparsity (mean gap 1.7) survival coverage is still %.3f-%.3f. Competing-risks CIFs and cause-specific effects stay at %.2f-%.2f (incl. with 30%% masked causes), smooth f(x) tracks the oracle, and results are insensitive to m in {5, 10, 20}.

2. **MI clearly beats the naive midpoint default** -- the comparison that matters for practice, since midpoint imputation is what people actually do. Under sparse inspection the paired coverage gain for S/hazard is %+.3f to %+.3f (paired MC SE <= %.3f), and midpoint is never significantly better on any estimand anywhere. The mechanism is fully diagnosed: midpoint discards imputation uncertainty (too-narrow CIs, Fig 4) around structurally biased point estimates; MI propagates the uncertainty and partly de-biases.

3. **The honest limitation: one-step MI under sparse inspection.** With mean inspection gaps of ~3.3, MI's point estimates inherit bias from the midpoint-initialized imputation fit, concentrated at early-to-mid t; SEs stay honest (be_coverage nominal, se_ratio ~ 1), so coverage *falls as n grows* (const-ph: %.3f at n=300 vs %.3f at n=1000) -- and Turnbull recovers S(1) on the same data. This motivates the iterated (refit-and-reimpute) extension scoped in the README; the per-(rep, method) seed substreams make it a drop-in addition over the same simulated data.

4. **External comparators behave exactly as theory says.** ic_par is nominal wherever its Weibull is correct (const cells: %.3f-%.3f at every rate, n) and fails under the peaked baseline (%.3f-%.3f) with no rescue from n -- the cost of parametric convenience. On S(t|x) point accuracy MI leads the non-oracle field (median RRMSE vs oracle: MI %.2f, midpoint %.2f, Turnbull %.2f; Turnbull beats MI in only %d/%d core cells): the Turnbull NPMLE remains the assumption-free point-only reference, but buys no smoothing, no covariate model and no CIs, while MI alone provides full regression inference.

5. **Practical guidance.** Under dense inspection (mean gap <= ~0.7 on this 10-unit horizon) pamm_ic inference can be used as-is for all estimands except the interval-average hazard near sharp peaks (a smoothing, not an IC, limitation -- the oracle hits the same ceiling). Already at moderate sparsity (mean gap ~1.7) the early-time bias begins to show: survival coverage dips to %.3f in the worst core cell (large n, bias-driven) even though effect estimates stay reliable. With sparser panels still, treat early-time S(t|x) point estimates as optimistic until the iterated-MI extension lands; effect estimates (beta, logHR(t), f(x)) remain trustworthy longer than survival curves do.

6. **Methodological note.** The three-gate review process earned its keep: the final gate caught an informative-censoring defect in the inspection DGP (decision log, 2026-06-11) precisely because the design included a *well-specified parametric comparator* whose coverage collapse could only mean a data-generation bug. Production was re-run on the corrected DGP; the pre-fix archive is retained for the paired comparison.\n",
  min(mi_dense$coverage), max(mi_dense$coverage),
  min(ph_mid$coverage), max(ph_mid$coverage),
  cr_rng[1], cr_rng[2],
  min(pd_sp$diff), max(pd_sp$diff), max(pd_sp$diff_mcse),
  mi_sp_n$coverage[mi_sp_n$n == 300], mi_sp_n$coverage[mi_sp_n$n == 1000],
  min(icp_const$coverage), max(icp_const$coverage),
  min(icp_peak$coverage), max(icp_peak$coverage),
  rrm("mi"), rrm("midpoint"), rrm("turnbull"), tb_v_mi$v, tb_v_mi$n,
  min(ph_mid$coverage)
))
```

# Appendix

## A1: Term-level summary tables

Full per-cell reference tables, one block per estimand: coverage (MC SE),
bias-eliminated coverage, bias (MC SE), RMSE (MC SE), avSE/empSE and median
CI width per (cell, method). Use them to look up any individual cell behind
the figures; the patterns are summarized in the Results section and the ANOVA
table above.

```{r tables, results = "asis"}
for (es in unique(tl$estimand)) {
  cat(sprintf("\n### %s\n\n", es))
  print(knitr::kable(
    tl |>
      filter(estimand == es) |>
      mutate(
        coverage = sprintf("%.3f (%.3f)", coverage, coverage_mcse),
        bias = sprintf("%.4f (%.4f)", bias, bias_mcse),
        rmse = sprintf("%.4f (%.4f)", rmse, rmse_mcse)
      ) |>
      select(cell_id, method, n_rep, coverage, be_coverage, bias, rmse,
             se_ratio, median_width, misspecified),
    digits = 3
  ))
}
```

## A2: Session / run metadata

```{r meta, results = "asis"}
cat(sprintf("- raw rows: %d\n", res$n_raw_rows))
cat(sprintf("- max reps/cell: %d\n", n_rep_prod))
cat(sprintf("- median fit times by method (s): %s\n",
  tl |> group_by(method) |> summarise(t = median(mean_fit_time, na.rm = TRUE)) |>
    (\(d) paste(sprintf("%s %.1f", d$method, d$t), collapse = ", "))()
))
```
