---
title: "Interval-censored PAMMs: Monte Carlo benchmark"
date: today
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 10
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(tibble)
})
bench_dir <- if (file.exists("config.R")) "." else "attic/ic-benchmark"
res <- readRDS(file.path(bench_dir, "results", "summary.rds"))
tl <- res$term_level
pw <- res$pointwise |>
  left_join(
    distinct(tl, cell_id, arm, baseline, effect, rate, n, mean_int),
    by = "cell_id"
  )
paired <- res$paired
theme_set(theme_bw(base_size = 12))
cols <- c(
  oracle = "#1b9e77", mi = "#1f78b4", midpoint = "#e31a1c",
  ic_par = "#ff7f00", ic_sp = "#6a3d9a", turnbull = "#b15928"
)
meth_lvls <- names(cols)
tl <- tl |> mutate(method = factor(method, levels = meth_lvls))
pw <- pw |> mutate(method = factor(method, levels = meth_lvls))
n_rep_prod <- max(tl$n_rep)
```

All prose numbers in this report are computed from `results/summary.rds`
(`r res$n_raw_rows` raw result rows, up to `r n_rep_prod` replications per
cell). Monte Carlo SEs for grid-averaged coverage are clustered at the
replication level. Cells where a method is misspecified by design are flagged
with open symbols.

# Design

## Aims

This benchmark validates and compares the interval-censored (IC) survival
workflow added to `pammtools`: `pamm_ic()` / `pamm_ic_cr()` fit piecewise
exponential additive mixed models (PAMMs) to interval-censored data by
multiple imputation — exact event times are drawn from a conditional
piecewise-exponential hazard initialized from a midpoint-imputed fit
(one-step MI), a PAMM is fit per imputation, and inference is pooled via
Rubin's rules. The study follows the ADEMP framework (Morris, White &
Crowther 2019, *Stat Med*); the full reviewed design is in `PLAN.md` and the
ADEMP document with its decision log in `README.md` (both in this folder).

- **A1 (validation):** does `pamm_ic` MI inference attain nominal 95%
  coverage and low bias for hazard, survival, covariate effects, time-varying
  effects and competing-risks CIFs, across inspection sparsity, inspection
  mechanisms and baseline-hazard shapes?
- **A2 (comparison):** how does it compare against established external IC
  methods — `icenReg::ic_par` (parametric Weibull PH), `icenReg::ic_sp`
  (semiparametric PH) and the Turnbull NPMLE (`icenReg::ic_np`) — including
  cells where the parametric comparator is **misspecified by design**? The
  headline misspecification exhibit is `ic_par`'s S(t|x) behavior under the
  non-Weibull "peaked" baseline; PH-only comparators in time-varying-effect
  cells are reported as *estimand mismatch* (they target a least-false
  constant logHR), never as headline "coverage failure".

The design was frozen pre-production at Gate R2 (three external review gates:
code, pilot, results; see the decision log in `README.md`). The earlier
200-rep prototype is `attic/ic-simulation-study.R`.

## Methods compared

| Method | Fit | Confidence intervals | Role |
|---|---|---|---|
| `oracle` | PAMM on the **exact** (uncensored) event times, analysis grid CUT | simulation CIs, 500 posterior draws | upper reference bound: best achievable on this grid; isolates the cost of interval censoring |
| `mi` | `pamm_ic` / `pamm_ic_cr`: m = 10 imputations (5/20 in spoke I), PAMM per imputation, Rubin-pooled | pooled posterior draws (500 total); scalar β via Rubin's rules with Barnard–Rubin small-sample df | the method under test |
| `midpoint` | single PAMM on midpoint-imputed times ((L + min(R, horizon))/2; right-censored at L if R = ∞) | simulation CIs, 500 draws | naive single-imputation comparator: ignores imputation uncertainty |
| `ic_par` | `icenReg::ic_par`, Weibull PH | Wald z for β; S/hazard via 500 normal coefficient draws pushed through the closed-form Weibull | external parametric comparator; **misspecified by design** under the peaked baseline and in TV cells |
| `ic_sp` | `icenReg::ic_sp`, semiparametric PH (NPMLE baseline) | β: bootstrap percentile (500 resamples) in 4 designated cells, point-only elsewhere; S(t\|x) point-only (the bootstrap covers coefficients, not the NPMLE baseline) | external semiparametric comparator |
| `turnbull` | `icenReg::ic_np` stratified by x (Turnbull NPMLE) | none (point accuracy only) | nonparametric accuracy benchmark for S(t\|x) in the simple cases |

All simulation-based CIs use the same number of draws (nsim = 500, equalized
so coverage differences are not tail-order-statistic artifacts; an LRZ
sensitivity run over nsim ∈ {200, 500, 1000} showed identical coverage).
Within each Turnbull equivalence class (where the NPMLE is flat or
undefined), S is scored at the midpoint of the upper/lower envelope and the
half-gap is recorded as `ambiguity`. Note one deliberate construction
difference: MI β CIs are Barnard–Rubin t, midpoint/oracle/ic_par β CIs are
Wald z. External methods apply only in the single-event, linear-effect cells;
the smooth-f(x) and competing-risks arms compare internal methods only
(mi/midpoint/oracle).

Model formulas (mgcv syntax): PH cells `~ s(tend, k = 10) + x`; TV cells add
`+ s(tend, by = x_num, k = 10)`; spoke S `~ s(tend, k = 10) + s(x, k = 10)`;
CR arm `~ s(tend, by = cause, k = 10) + cause + cause:x`.

## Data-generating processes and design

Data are simulated with `pammtools::sim_pexp()` on a fine grid
(FINE = seq(0, 10, 0.05)) and interval-censored with
`pammtools::add_inspections()`. All methods analyze the same frozen grid
CUT = seq(0, 10, 0.5); estimands are evaluated at t ∈ {1, 2, 3, 5, 7}.

**Hazards.** Log-baseline shapes: *const* −1.5 (Weibull-compatible; `ic_par`
correctly specified), *peaked* −3.2 + 1.6·exp(−(t−3)²/2) (non-monotone; no
Weibull can match it), *rising* −3.7 + 0.12t (spoke R). Covariate effects:
*ph* β·x with β = 0.6 and binary x (p = 0.5 — chosen so the stratified
Turnbull NPMLE is directly comparable); *tv* β(t)·x with
β(t) = 1.2·exp(−0.3t); *smooth* f(x) = 0.8·sin(πx) with x ~ U(−1, 1). The CR
arm uses two independent latent cause-specific hazards (cause 1 peaked with
β₁ = 0.7, cause 2 rising with β₂ = −0.5, x ~ U(−1, 1)); the `cr_unknown` cell
additionally masks 30% of observed event causes to exercise
`pamm_ic_cr`'s cause-imputation branch.

**Inspection mechanisms.** *random*: per-subject inspection grids with
Exp(rate) gaps, rate ∈ {1.5, 0.6, 0.3} — mean gap 1/rate ≈ 0.67 / 1.67 / 3.33,
i.e. dense → sparse interval censoring. *fixed*: one shared deterministic
schedule for all subjects (dense gap 2/3, sparse gap 10/3). *mixed*: the fixed
schedule shifted per subject by U(0, gap). The event time is bracketed by the
nearest inspections (L, R]; events after the last inspection or horizon are
right-censored. **Caveat (fixed schedule):** with one shared grid, S(t) is
nonparametrically identified *only at the schedule points*; elsewhere all
methods score model-based interpolation, and sparse-grid hazard estimands can
be degenerate. These cells are robustness exhibits, never headlined alongside
the random-inspection cells.

**Truth computation.** `sim_pexp()` draws from a *left-endpoint
piecewise-constant* hazard on FINE, not from the smooth formula. All truths
(S, interval-average hazard, logHR, CIF) are therefore computed **exactly**
from those same FINE rate vectors via cumulative sums (`dgp.R`) — truth and
simulator derive from one definition and cannot diverge. Phase 0 verified
this against a large-sample Kaplan–Meier (max |ΔS| ≤ 0.0016) and an
independent `survival::survfit` Turnbull cross-check.

**Design.** A core factorial plus targeted spokes; the cells present in this
run:

```{r design-table}
design <- tl |>
  distinct(cell_id, arm, baseline, effect, mechanism, rate, n, m) |>
  group_by(arm) |>
  summarise(
    cells = n(),
    baseline = paste(sort(unique(baseline)), collapse = ", "),
    effect = paste(sort(unique(effect)), collapse = ", "),
    mechanism = paste(sort(unique(mechanism)), collapse = ", "),
    rate = paste(sort(unique(rate), decreasing = TRUE), collapse = ", "),
    n = paste(sort(unique(n)), collapse = ", "),
    m = paste(sort(unique(m)), collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(match(arm, c(
    "core", "mech", "rising", "smooth", "mimp", "cr", "cr_unknown"
  )))
knitr::kable(design)
```

The core arm crosses baseline × effect × inspection rate × n under random
inspections; the spokes vary one factor each off the (peaked, ph, n = 300)
reference cell: inspection *mech*anism, *rising* baseline, *smooth* f(x),
number of imputations (*mimp*), and competing risks. Acknowledged gaps:
mechanism × n and mechanism × effect interactions are untested, so spoke
conclusions are conditional on the reference cell.

## Estimands

All estimands are aligned exactly to the piecewise-exponential DGP:

- **S(t|x)** at TEVAL × x ∈ {0, 1} — well-defined for every method, hence the
  headline estimand for cross-method comparison.
- **Hazard** as the *interval average* h̄ⱼ = (H(tⱼ) − H(tⱼ₋₁)) / 0.5 on the CUT
  interval ending at each TEVAL. Rationale: a PAMM estimates piecewise-constant
  hazards on CUT; scoring it against the smooth endpoint hazard h(tⱼ) would
  build up to ~25% "bias" on the peak flanks for *all* methods — a
  discretization artifact, not a method property. `ic_par` is scored the same
  way via closed-form cumulative-hazard differences.
- **logHR(t)** = log h̄ⱼ(1) − log h̄ⱼ(0); constant = β under PH. In TV cells the
  PH-only comparators are evaluated against the *least-false constant logHR*
  (Cox projection from one large-n exact-time fit; overlay in Fig 5) rather
  than as covering β(t).
- **β** (scalar, PH cells); **f(x)** on an x-grid, with estimate and truth both
  centered to mean zero over the grid (spoke S); **CIF₁/CIF₂(t)** at x = 0
  (CR arm).

## Performance measures and Monte Carlo precision

Per Morris et al. (2019): bias, empirical SE, RMSE, 95% CI coverage,
bias-eliminated coverage (CI shifted to cover the mean estimate — separates
"CIs too narrow" from "point estimate biased"), CI width, the avSE/empSE
ratio (model-based vs empirical SE; the most diagnostic quantity behind
coverage failures), failure rates and fit times — each with Monte Carlo SEs.
Two aggregation details matter:

- avSE/empSE for grid-valued estimands is computed *within* each (t, x) grid
  point and then median-summarized; pooling errors across the grid mixes
  cross-point bias heterogeneity into "empSE" and invalidates the ratio.
- CI widths are summarized by medians and q90 with a flag for extreme widths,
  since single pathological replications (vacuous CIs in weakly identified
  cells) make raw means meaningless.

n_sim = 500 reps/cell gives a coverage MC SE of ≈ 1.0% at 95% nominal. The
headline MI-vs-midpoint contrasts therefore use **paired differences** (both
methods see the same data in each rep) with paired MC SEs, which are
substantially more precise than per-method coverage; term-level
(grid-averaged) coverage MC SEs are clustered at the replication level.

# Results

## Failure rates

```{r failures, results = "asis"}
fl <- res$failures |> filter(n_fail > 0)
if (nrow(fl) == 0) {
  cat("No method failures occurred in any cell.\n")
} else {
  cat(sprintf(
    "%d (cell, method) combinations had failures; max failure rate %.1f%% (%s, %s).\n\n",
    nrow(fl), 100 * max(fl$fail_rate),
    fl$cell_id[which.max(fl$fail_rate)], fl$method[which.max(fl$fail_rate)]
  ))
  knitr::kable(fl, digits = 3)
}
if (nrow(res$missing) > 0) {
  cat(sprintf("\n**WARNING: %d missing (cell, rep) combinations.**\n",
              nrow(res$missing)))
}
```

## Fig 1 (headline): coverage of S(t|x) and interval-average hazard

```{r fig1, fig-height: 7}
d1 <- tl |>
  filter(arm == "core", estimand %in% c("surv", "hazard"), !is.na(coverage))
ggplot(d1, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  geom_line(alpha = .5) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage (grid-avg)",
    colour = NULL, shape = "misspecified",
    title = "Coverage of S(t|x) and interval-average hazard, core factorial",
    subtitle = "pointranges: +-1 MC SE (rep-clustered)"
  )
```

```{r fig1-prose, results = "asis"}
mi_s <- d1 |> filter(method == "mi", estimand == "surv")
md_s <- d1 |> filter(method == "midpoint", estimand == "surv")
cat(sprintf(
  "MI survival coverage across core cells: %.3f-%.3f (median %.3f). Midpoint: %.3f-%.3f (median %.3f).\n",
  min(mi_s$coverage), max(mi_s$coverage), median(mi_s$coverage),
  min(md_s$coverage), max(md_s$coverage), median(md_s$coverage)
))
```

## Fig 2: beta / logHR(t) coverage + paired MI−midpoint differences

```{r fig2}
d2 <- tl |>
  filter(arm == "core", estimand %in% c("beta", "logHR"), !is.na(coverage))
ggplot(d2, aes(mean_int, coverage, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.12)) +
  facet_grid(estimand + n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Covariate-effect coverage (beta in PH cells, logHR(t) grid-avg)",
    subtitle = "open symbols: misspecified by design (constant-beta methods vs beta(t))"
  )
```

```{r fig2b, fig-height: 4}
ggplot(
  paired |> filter(arm == "core"),
  aes(factor(rate, levels = c(1.5, 0.6, 0.3)), diff, colour = estimand)
) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(ymin = diff - diff_mcse, ymax = diff + diff_mcse),
    position = position_dodge(width = 0.4)
  ) +
  facet_grid(n ~ baseline + effect) +
  labs(
    x = "inspection rate (dense -> sparse)", y = "coverage(MI) - coverage(midpoint)",
    title = "Paired MI - midpoint coverage differences (same data per rep)",
    subtitle = "pointranges: +-1 paired MC SE"
  )
```

## Fig 3: point accuracy (RMSE, incl. Turnbull)

```{r fig3}
d3 <- tl |> filter(arm == "core", estimand == "surv")
base_rmse <- d3 |>
  filter(method == "oracle") |>
  select(cell_id, rmse_oracle = rmse)
d3 <- left_join(d3, base_rmse, by = "cell_id") |>
  mutate(rrmse = rmse / rmse_oracle)
ggplot(d3, aes(mean_int, rrmse, colour = method, shape = misspecified)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(n ~ baseline + effect) +
  scale_colour_manual(values = cols) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1)) +
  scale_y_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "mean inspection-interval length", y = "RMSE relative to oracle (log)",
    title = "S(t|x) point-estimate accuracy relative to the oracle"
  )
```

```{r fig3-prose, results = "asis"}
tb <- tl |> filter(method == "turnbull", estimand == "surv", arm == "core")
if (nrow(tb)) {
  rr <- d3 |>
    filter(method %in% c("mi", "midpoint", "turnbull")) |>
    group_by(method) |>
    summarise(med_rrmse = median(rrmse, na.rm = TRUE), .groups = "drop")
  n_tb_beats_mi <- d3 |>
    select(cell_id, method, rrmse) |>
    pivot_wider(names_from = method, values_from = rrmse) |>
    summarise(v = sum(turnbull < mi, na.rm = TRUE), n = sum(!is.na(turnbull)))
  cat(sprintf(
    "Median RRMSE vs oracle: MI %.2f, midpoint %.2f, Turnbull %.2f; Turnbull beats MI in %d/%d core cells (it is a step-function NPMLE without smoothing). Turnbull mean equivalence-class half-gap (scoring ambiguity) ranges %.3f-%.3f across core cells.\n",
    rr$med_rrmse[rr$method == "mi"], rr$med_rrmse[rr$method == "midpoint"],
    rr$med_rrmse[rr$method == "turnbull"],
    n_tb_beats_mi$v, n_tb_beats_mi$n,
    min(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE),
    max(pw$mean_ambiguity[pw$method == "turnbull"], na.rm = TRUE)
  ))
}
```

## Fig 4: CI width and avSE/empSE (the mechanism)

```{r fig4}
d4 <- tl |>
  filter(arm == "core", estimand == "surv",
         method %in% c("mi", "midpoint", "oracle", "ic_par")) |>
  pivot_longer(c(median_width, se_ratio), names_to = "what")
ggplot(d4, aes(mean_int, value, colour = method)) +
  geom_point(size = 2) + geom_line(alpha = .5) +
  facet_grid(what ~ baseline + effect + n, scales = "free_y") +
  scale_colour_manual(values = cols) +
  labs(
    x = "mean inspection-interval length", y = NULL,
    title = "S(t|x): median CI width and avSE/empSE ratio (median over grid)",
    subtitle = "se_ratio < 1: model-based SE understates sampling variability"
  )
```

## Fig 5: time-varying effect recovery

```{r fig5}
lf_file <- file.path(bench_dir, "results", "least-false.rds")
d5 <- pw |>
  filter(effect == "tv", estimand == "logHR", n == 300,
         method %in% c("mi", "midpoint", "oracle", "ic_par"))
if (nrow(d5) > 0) {
  lf_lines <- if (file.exists(lf_file)) {
    readRDS(lf_file) |>
      distinct(baseline, lf_cox) |>
      semi_join(d5, by = "baseline")
  } else {
    NULL
  }
  p5 <- ggplot(d5, aes(t, mean_est, colour = method)) +
    geom_line(aes(y = truth), colour = "black", linewidth = 0.9) +
    geom_pointrange(
      aes(ymin = mean_est - bias_mcse, ymax = mean_est + bias_mcse),
      position = position_dodge(width = 0.3)
    ) +
    geom_line(alpha = 0.5, position = position_dodge(width = 0.3)) +
    facet_grid(baseline ~ rate) +
    scale_colour_manual(values = cols) +
    labs(
      x = "t", y = "logHR(t)",
      title = "TV cells: mean estimated logHR(t) vs truth (black), n = 300",
      subtitle = "dashed grey: least-false constant (Cox projection) -- ic_par's actual target"
    )
  if (!is.null(lf_lines)) {
    p5 <- p5 +
      geom_hline(
        data = lf_lines, aes(yintercept = lf_cox),
        linetype = 2, colour = "grey50"
      )
  }
  print(p5)
} else {
  cat("(no TV-cell data in this run)")
}
```

```{r fig5-prose, results = "asis"}
if (file.exists(lf_file)) {
  lf <- readRDS(lf_file) |> distinct(baseline, lf_cox, lf_weibull)
  for (i in seq_len(nrow(lf))) {
    cat(sprintf(
      "- %s baseline: least-false constant logHR (exact-time convention): Cox %.3f, Weibull %.3f\n",
      lf$baseline[i], lf$lf_cox[i], lf$lf_weibull[i]
    ))
  }
} else {
  cat("Least-false projections not yet computed (results/least-false.rds missing).\n")
}
```

## Fig 6: competing risks, smooth f(x), m-sensitivity

```{r fig6a}
d6 <- tl |> filter(arm %in% c("cr", "cr_unknown"), !is.na(coverage))
ggplot(d6, aes(mean_int, coverage, colour = method, shape = arm)) +
  geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(
    ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
  ), position = position_dodge(width = 0.1)) +
  facet_wrap(~estimand, nrow = 1) +
  scale_colour_manual(values = cols) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "mean inspection-interval length", y = "95% CI coverage",
    title = "Competing-risks arm (triangle: 30% unknown causes)"
  )
```

```{r fig6b, fig-height: 4}
d6b <- tl |> filter(arm == "smooth", estimand == "f_x")
if (nrow(d6b)) {
  ggplot(d6b, aes(factor(rate), coverage, colour = method)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    scale_colour_manual(values = cols) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = "inspection rate", y = "95% CI coverage",
      title = "Smooth f(x) coverage (spoke S, grid-avg over x)"
    )
}
```

```{r fig6c, fig-height: 4}
d6c <- tl |>
  filter(arm == "mimp" | cell_id == "core-peaked-ph-random-r0.6-n300-m10",
         method == "mi", estimand %in% c("surv", "hazard", "beta"))
if (nrow(d6c)) {
  ggplot(d6c, aes(factor(m), coverage, colour = estimand)) +
    geom_hline(yintercept = 0.95, linetype = 2, colour = "grey40") +
    geom_pointrange(aes(
      ymin = coverage - coverage_mcse, ymax = coverage + coverage_mcse
    ), position = position_dodge(width = 0.3)) +
    coord_cartesian(ylim = c(0.7, 1)) +
    labs(
      x = "number of imputations m", y = "95% CI coverage",
      title = "m-sensitivity (spoke I; peaked, PH, rate 0.6, n = 300)"
    )
}
```

## Term-level summary tables

```{r tables, results = "asis"}
for (es in unique(tl$estimand)) {
  cat(sprintf("\n### %s\n\n", es))
  print(knitr::kable(
    tl |>
      filter(estimand == es) |>
      mutate(
        coverage = sprintf("%.3f (%.3f)", coverage, coverage_mcse),
        bias = sprintf("%.4f (%.4f)", bias, bias_mcse),
        rmse = sprintf("%.4f (%.4f)", rmse, rmse_mcse)
      ) |>
      select(cell_id, method, n_rep, coverage, be_coverage, bias, rmse,
             se_ratio, median_width, misspecified),
    digits = 3
  ))
}
```

## Session / run metadata

```{r meta, results = "asis"}
cat(sprintf("- raw rows: %d\n", res$n_raw_rows))
cat(sprintf("- max reps/cell: %d\n", n_rep_prod))
cat(sprintf("- median fit times by method (s): %s\n",
  tl |> group_by(method) |> summarise(t = median(mean_fit_time, na.rm = TRUE)) |>
    (\(d) paste(sprintf("%s %.1f", d$method, d$t), collapse = ", "))()
))
```
