---
title: "Interval-Censored Data"
author: "pammtools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Remote.bib
vignette: >
  %\VignetteIndexEntry{Interval-Censored Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.height = 4,
  fig.width  = 6
)
```

```{r, message = FALSE}
library(pammtools)
library(survival)
library(dplyr)
library(ggplot2)
```

## Background

In many studies the event time is not observed exactly but only known to lie in
an interval $(L_i, R_i]$ -- for example because subjects are examined at a
sequence of *inspection times* and we only learn that the event happened between
the last clean and the first positive visit. This is **interval censoring**.

A piecewise-exponential additive model (PAMM) assumes a piecewise-constant
hazard, so the conditional distribution of the event time within $(L_i, R_i]$ is
known in closed form. `pammtools` exploits this with a **multiple imputation
(MI)** strategy [@pan2000]: exact event times are
repeatedly drawn from $p(T \mid L_i < T \le R_i, x_i, \theta)$, each completed
data set is transformed and re-fit with the *standard* right-censored pipeline,
and the fits are pooled for inference following Rubin's rules [@rubin1987].
Naive midpoint imputation, by contrast, typically yields reasonable point
estimates but invalid (too narrow) confidence intervals. Related model-based
imputation schemes for the interval-censored hazard are discussed by
@bebchuk2000; the `icenReg` package [@andersonbergman2017] provides
non-parametric and parametric alternatives that are useful as a cross-check.

Interval-censored data are specified through the usual **survival** interface,
`Surv(L, R, type = "interval2")`, where `L == R` denotes an exact time,
`R == Inf` a right-censored observation, `L == 0` a left-censored one, and
`0 < L < R < Inf` a genuine interval.

## Simulating interval-censored data

`add_inspections()` turns exact simulated times (here from `sim_pexp()`) into
interval-censored panel data by drawing (per-subject, by default) inspection times. The true
event time is retained in `true_time` so we can check our work.

Here a binary covariate `trt` acts through a *time-varying* effect that is
**absent at baseline and grows over follow-up**. The control group (`trt == 0`)
has a *constant* hazard; the other group starts at the same hazard but its
log-hazard then rises linearly in time -- the log-hazard ratio is `0.15 * t`,
zero at `t = 0` -- so the two hazards start together and fan apart, an
increasingly harmful effect and a clear departure from proportional hazards.
Because factors enter the (numeric) linear predictor directly, the effect is
written with an explicit indicator `(trt == 1)`.

The inspection `rate` controls how much information the interval censoring
destroys: a dense grid pins each event down tightly and leaves little to impute,
a sparse one brackets events only loosely. We use a moderate schedule
(`rate = 1.5`, mean inter-inspection gap $2/3$) -- dense enough that these simple
hazards are recovered well, yet still leaving each event bracketed in an interval
that straddles several analysis cut-points, so the gap between proper MI and
naive midpoint intervals stays appreciable.

```{r}
set.seed(3)
df  <- data.frame(trt = rbinom(1000, 1, 0.5) |> as.factor())
sdf <- sim_pexp(~ -2.2 + 0.15 * t * (trt == 1), df,
                cut = seq(0, 10, by = 0.1))
icd <- add_inspections(sdf, rate = 1.5, max_time = 10)
icd %>% select(id, trt, true_time, L, R) %>% head()
```

## Fitting a PAMM by multiple imputation

`pamm_ic()` runs the whole MI workflow: an initial midpoint fit, then `m`
rounds of (proper) imputation and re-fitting on a fixed interval grid.

We give each group its own smooth log-baseline-hazard via `s(tend, by = trt)`
(a factor `by`-variable yields one centred smooth per level) plus the parametric
`trt` main effect for the overall level difference between groups.

```{r}
fit <- pamm_ic(
  Surv(L, R, type = "interval2") ~ trt,
  data          = icd,
  model_formula = ped_status ~ trt + s(tend, by = trt),
  cut           = seq(0, 10, by = 0.5),
  m             = 20
)
fit
```

`print()` gives a compact overview; `summary()` reports the **pooled** inference:
parametric coefficients and covariances are combined with Rubin's rules (so the
reported standard errors include the between-imputation variability), term
p-values use the median-p rule [@bolt2022], and the fraction of missing
information induced by the interval censoring is reported as well.

```{r}
summary(fit)
```

The `m` imputation fits are stored in `fit$fits` in a slimmed form (the
per-observation slots are dropped, so memory does not grow with the data set
size), and pooled summary diagnostics live in `fit$pooled`. Because smooth bases
can differ across imputed data sets, `fit$pooled` is not a single `gam` object
for direct `predict()` or `plot()` calls. The pooled `add_*()` methods below
operate on the imputation fits directly: they combine the per-imputation
posterior draws so the intervals reflect both within- and between-imputation
uncertainty.

We compare two PAMM estimates per group. The **MI** estimate pools the imputation
fits, so its intervals carry both the within-fit and the between-imputation
uncertainty. The **midpoint** estimate is the initialiser fit (`fit$init_fit`):
a single PAMM fit to the interval midpoints, which treats those imputed times as
if observed. We request **simulation-based** intervals (`ci_type = "sim"`) for the
midpoint fit so that both estimates use the same (proper) interval machinery as
the pooled `pamm_ic` methods -- the pooled methods always simulate from each fit's
posterior, so this is a like-for-like comparison that differs only by the
between-imputation variance. (The default `ci_type = "default"` instead cumulates
the pointwise hazard limits, which yields a noticeably more conservative band.)

```{r}
ped <- as_ped(icd, Surv(L, R, type = "interval2") ~ trt, cut = seq(0, 10, by = 0.5))
nd  <- make_newdata(ped, tend = unique(tend), trt = unique(trt)) %>% group_by(trt)

# midpoint = initialiser fit (single completed data set, no imputation uncertainty)
set.seed(1)   # the simulation-based intervals draw from each fit's posterior
surv_mid   <- add_surv_prob(nd, fit$init_fit, ci_type = "sim", nsim = 1000)
surv_mi    <- add_surv_prob(nd, fit, nsim = 1000)
hazard_mid <- add_hazard(nd, fit$init_fit, ci_type = "sim", nsim = 1000)
hazard_mi  <- add_hazard(nd, fit, nsim = 1000)
```

As a non-parametric cross-check we add the **Turnbull** estimator -- the NPMLE of
the survival function for interval-censored data -- computed per group.
`survival::survfit()` runs the Turnbull EM algorithm when handed an `interval2`
response and returns pointwise confidence limits.

```{r}
tb    <- survfit(Surv(L, R, type = "interval2") ~ trt, data = icd)
tb_df <- data.frame(
  tend  = tb$time,
  surv  = tb$surv,
  lower = tb$lower,
  upper = tb$upper,
  trt   = factor(rep(c(0, 1), tb$strata))   # strata are ordered trt = 0, then 1
)
```

All three recover the diverging survival curves well -- the groups start together
and the increasingly harmful effect opens a growing gap. The midpoint intervals
are the tightest -- they ignore the imputation uncertainty and are therefore
over-confident; the MI intervals are wider (the censoring here induces a sizeable
fraction of missing information, see `summary(fit)` above), and the
non-parametric Turnbull intervals are wider and more jagged still.

```{r, fig.width = 9, fig.height = 3.6}
prep <- function(d, method) {
  transmute(d, tend, trt = factor(trt),
    surv = surv_prob, lower = surv_lower, upper = surv_upper, method = method)
}
surv_all <- bind_rows(
  prep(surv_mid, "PAMM (midpoint)"),
  prep(surv_mi,  "PAMM (MI)"),
  transmute(tb_df, tend, trt, surv, lower, upper, method = "Turnbull"))
surv_all$method <- factor(surv_all$method,
  levels = c("PAMM (midpoint)", "PAMM (MI)", "Turnbull"))

ggplot(surv_all, aes(x = tend, y = surv, color = trt, fill = trt)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line() +
  facet_grid(~ method) +
  scale_color_viridis_d(end = 0.8, aesthetics = c("colour", "fill")) +
  labs(x = "time", y = "S(t)", color = "trt", fill = "trt",
    title = "Interval-censored S(t): midpoint vs. multiple imputation vs. Turnbull")
```

The over-confidence of the midpoint fit is easy to quantify -- the width 
of the pointwise survival intervals is always wider for MI than fior midpoint 
imputation:

```{r}
# midpoint-CI-width / MI-CI-width
summary((surv_mi$surv_upper  - surv_mi$surv_lower) / 
        (surv_mid$surv_upper - surv_mid$surv_lower))
```

The effect is more visible on the **hazard** scale, which depends directly on
exact event timing which interval censoring leaves more uncertain. Overlaying
the two methods within each group shows the MI hazard bands enveloping
the (over-confident) midpoint-based bands, while the point estimates mostly agree:

```{r, fig.width = 8, fig.height = 4}
haz_all <- bind_rows(
  mutate(hazard_mid, method = "PAMM (midpoint)"),
  mutate(hazard_mi,  method = "PAMM (MI)")) %>%
  mutate(trt = paste0("trt = ", trt),
    method = factor(method, levels = c("PAMM (midpoint)", "PAMM (MI)")))

# true conditional hazards from the data-generating log-hazard
tgrid <- seq(0, 10, by = 0.1)
haz_true <- bind_rows(
  data.frame(tend = tgrid, trt = "trt = 0", hazard = exp(-2.2)),
  data.frame(tend = tgrid, trt = "trt = 1", hazard = exp(-2.2 + 0.15 * tgrid)))

ggplot(haz_all, aes(x = tend, y = hazard, color = method, fill = method)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.3, color = NA) +
  geom_step() +
  geom_line(data = haz_true, aes(x = tend, y = hazard), inherit.aes = FALSE,
    color = "gold", alpha = 0.5, linewidth = 0.9, linetype = 2) +
  facet_wrap(~ trt) +
  scale_color_viridis_d(end = 0.5, aesthetics = c("colour", "fill"), option = "C") +
  labs(x = "time", y = "hazard rate", color = NULL, fill = NULL,
    title = "Conditional hazard rate: midpoint vs. multiple imputation\n (true rates in gold)")
```

The true rates (gold) are tracked closely across the whole range -- the constant
`trt = 0` hazard and the rising `trt = 1` hazard are both recovered well. The
midpoint and MI point estimates nearly coincide, consistent with midpoint giving
reasonable point estimates but over-confident intervals; the MI bands are visibly
wider throughout.

Averaged over the grid, the MI hazard intervals are appreciably wider:

```{r}
c(midpoint = mean(hazard_mid$ci_upper - hazard_mid$ci_lower),
  MI       = mean(hazard_mi$ci_upper  - hazard_mi$ci_lower)) %>% round(3)
```

So the imputation uncertainty inflates the hazard intervals more than the
survival intervals: $S(t)$ aggregates the hazard over time and the group-level
survival stays fairly well identified, whereas the hazard itself is tied directly
to exact event times, which interval censoring leaves uncertain (cf.\ the
large fraction of missing information reported by `summary()` above).  
The same machinery is available for `add_cumu_hazard()`.

## Competing risks

With $K$ competing causes, `pamm_ic_cr()` imputes both the event time and the
cause from a fitted **cause-specific** PAMM, and re-fits cause-specific hazards;
cumulative incidence functions are then obtained from those hazards with
`add_cif()`. Note that this targets cause-specific hazards (and CIFs derived from
them), and so differs from @delord2016, who develop MI for interval-censored
competing risks in the **subdistribution-hazard** (Fine--Gray) framework; the
scheme here is the cause-specific analogue. Because we impute from a
cause-specific model and analyse with one, the imputation and analysis models are
congenial.

```{r, eval = FALSE}
# WIP -cont here
icd$cause <- with(icd, ifelse(true_time <= 10,
  sample(c(1, 2), nrow(icd), TRUE, c(.6, .4)), 0))
fit_cr <- pamm_ic_cr(
  Surv(L, R, type = "interval2") ~ trt * cause, # linear TVEs for trt
  cause = "cause", cut = seq(0, 10, by = 0.5),
  data = icd, m = 10)
summary(fit_cr)

ped_cr <- as_ped(
  transform(icd, time = pmin(true_time, 10), status = cause),
  Surv(time, status) ~ trt * cause, 
  cut = seq(0, 10, by = 0.5))
nd_cr  <- make_newdata(ped_cr, 
                       tend = unique(tend), cause = unique(cause), trt = unique(trt)) |>
  group_by(cause)
add_cif(nd_cr, fit_cr) |> 
  ggplot(aes(x = tend, y = cif, fill = cause, col = cause)) + 
  geom_area(position = "stack") + 
  scale_color_viridis_d() +
  facet_grid(~trt)
```

Neither treatment effect nor baseline hazards are different across causes in this simple example, and this is recovered well in the estimated CIFs.

### How the time and cause are imputed

For a subject known to fail from cause $k$ in $(L, R]$, the correct conditional
density of the event time is the cause-$k$ *sub-density*, restricted to the
interval,
$$
p(t \mid L < T \le R,\ \text{cause}=k) \;\propto\; f_k(t) = h_k(t)\,S(t),
\qquad
S(t) = \exp\!\Big(-\textstyle\int_0^t h_\bullet(u)\,du\Big),
$$
where $h_\bullet = \sum_j h_j$ is the **all-cause** hazard. The all-cause hazard
is unavoidable here: to fail from cause $k$ *at* $t$, the subject must first
survive *all* causes up to $t$ (hence the all-cause survival $S$) and then fail
from cause $k$ (hence $h_k$). Drawing instead from a "cause-$k$-only" law
$h_k(t)\exp(-\int h_k)$ would be the classic net/latent-time error -- it ignores
that a competing cause may remove the subject before $t$, biasing imputed times
too late.

The all-cause *conditional* CDF inverts in closed form on the piecewise grid
(it is just the single-event sampler applied to $S = \exp(-H_\bullet)$), whereas
$f_k(t) = h_k(t)S(t)$ has no equally clean inverse. So `pamm_ic_cr()` uses the
all-cause conditional only as an easy-to-invert **proposal** and then **accepts**
the draw with probability $h_k(t)/h_\bullet(t) \le 1$. The accepted draws are
proportional to (proposal) $\times$ (acceptance) $= h_\bullet(t)S(t)\cdot
h_k(t)/h_\bullet(t) \propto h_k(t)S(t) = f_k(t)$ -- exactly the target; the
$h_\bullet$ cancels. For an **unknown** cause the same machinery is read the
other way: the time is drawn from the all-cause conditional and the cause is
then sampled with probability $h_k(t)/h_\bullet(t)$, the same
`hazard / total_hazard` ratio used to build CIF increments in `add_cif()`.

Two consequences are worth noting:

* If the cause-specific hazards are proportional ($h_k/h_\bullet$ constant over
  time) the acceptance is uniform and the all-cause conditional already *equals*
  the cause-$k$ conditional -- the rejection step is a no-op. It only matters
  when the cause mix shifts over follow-up (e.g. one cause dominates early,
  another late), which is exactly when cause-$k$ times need to be re-weighted.
* For a **rare** cause $h_k/h_\bullet$ is small everywhere, so acceptance is low;
  the sampler caps the number of tries and falls back to the last proposal, which
  can introduce a slight bias. For strongly imbalanced causes a direct inversion
  of the cause-$k$ sub-cumulative would be more robust -- a possible refinement.

## Notes and limitations

* The interval cut-points are fixed once and shared across all imputations; this
  is required for the imputations to be poolable.
* `proper = TRUE` (the default) draws the imputation-model coefficients from
  their posterior before each imputation; this is what makes the pooled
  intervals calibrated. `proper = FALSE` is for diagnostics only.
* Recurrent-event and general multi-state interval censoring are not yet
  supported: imputing one transition time changes the at-risk window for the
  next, so the imputations must be drawn jointly along each subject's path -- a
  planned extension.

## References
