---
title: "Interval-Censored Data"
author: "pammtools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Remote.bib
vignette: >
  %\VignetteIndexEntry{Interval-Censored Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.height = 4,
  fig.width  = 6
)
```

```{r, message = FALSE}
library(pammtools)
library(dplyr)
library(ggplot2)
```

## Background

In many studies the event time is not observed exactly but only known to lie in
an interval $(L_i, R_i]$ -- for example because subjects are examined at a
sequence of *inspection times* and we only learn that the event happened between
the last clean and the first positive visit. This is **interval censoring**.

A piecewise-exponential additive model (PAMM) assumes a piecewise-constant
hazard, so the conditional distribution of the event time within $(L_i, R_i]$ is
known in closed form. `pammtools` exploits this with a **multiple imputation
(MI)** strategy [@pan2000; @delord2016]: exact event times are
repeatedly drawn from $p(T \mid L_i < T \le R_i, x_i, \theta)$, each completed
data set is transformed and re-fit with the *standard* right-censored pipeline,
and the fits are pooled for inference following Rubin's rules [@rubin1987].
Naive midpoint imputation, by contrast, typically yields reasonable point
estimates but invalid (too narrow) confidence intervals. Related model-based
imputation schemes for the interval-censored hazard are discussed by
@bebchuk2000; the `icenReg` package [@andersonbergman2017] provides
non-parametric and parametric alternatives that are useful as a cross-check.

Interval-censored data are specified through the usual **survival** interface,
`Surv(L, R, type = "interval2")`, where `L == R` denotes an exact time,
`R == Inf` a right-censored observation, `L == 0` a left-censored one, and
`0 < L < R < Inf` a genuine interval.

## Simulating interval-censored data

`add_inspections()` turns exact simulated times (here from `sim_pexp()`) into
interval-censored panel data by drawing per-subject inspection times. The true
event time is retained in `true_time` so we can check our work.

```{r}
set.seed(2024)
df  <- data.frame(x = runif(300, -1, 1))
sdf <- sim_pexp(~ -2 + 0.5 * x, df, cut = seq(0, 10, by = 0.1))
icd <- add_inspections(sdf, rate = 0.75, max_time = 10)
icd %>% select(id, x, true_time, L, R) %>% head()
```

## Fitting a PAMM by multiple imputation

`pamm_ic()` runs the whole MI workflow: an initial midpoint fit, then `m`
rounds of (proper) imputation and re-fitting on a fixed interval grid.

```{r}
fit <- pamm_ic(
  Surv(L, R, type = "interval2") ~ x,
  data = icd,
  cut  = seq(0, 10, by = 0.5),
  m    = 10
)
fit
```

Each element of `fit$fits` is an ordinary `pamm` object, but the pooled
`add_*()` methods do the right thing automatically: they combine the
per-imputation posterior draws so the intervals reflect both within- and
between-imputation uncertainty.

```{r}
ped <- as_ped(icd, Surv(L, R, type = "interval2") ~ x, cut = seq(0, 10, by = 0.5))
nd  <- make_newdata(ped, tend = unique(tend))

surv <- add_surv_prob(nd, fit)
ggplot(surv, aes(x = tend, y = surv_prob)) +
  geom_step() +
  geom_stepribbon(aes(ymin = surv_lower, ymax = surv_upper), alpha = 0.2) +
  labs(x = "time", y = "S(t)", title = "Pooled survival probability (MI)")
```

The same applies to `add_hazard()` and `add_cumu_hazard()`.

## Competing risks

When several competing causes are present, `pamm_ic_cr()` draws the event time
from the all-cause conditional hazard and assigns a cause (observed causes are
retained; unknown causes are sampled in proportion to the cause-specific
hazards). Cumulative incidence functions are obtained with `add_cif()`.

```{r, eval = FALSE}
icd$cause <- with(icd, ifelse(true_time <= 10,
  sample(c(1, 2), nrow(icd), TRUE, c(.6, .4)), 0))
fit_cr <- pamm_ic_cr(Surv(L, R, type = "interval2") ~ x, icd,
  cause = "cause", cut = seq(0, 10, by = 0.5), m = 10)
nd_cr  <- make_newdata(ped_cr, tend = unique(tend), cause = unique(cause)) |>
  group_by(cause)
add_cif(nd_cr, fit_cr)
```

## Notes and limitations

* The interval cut-points are fixed once and shared across all imputations; this
  is required for the imputations to be poolable.
* `proper = TRUE` (the default) draws the imputation-model coefficients from
  their posterior before each imputation; this is what makes the pooled
  intervals calibrated. `proper = FALSE` is for diagnostics only.
* Recurrent-event and general multi-state interval censoring are not yet
  supported: imputing one transition time changes the at-risk window for the
  next, so the imputations must be drawn jointly along each subject's path -- a
  planned extension.

## References
